Teaching Applied Functional Programming\\ \small{Experiences from the functional programming course at TU Delft}

### 2013-10-22 latex/paper.tex
Update for ICSE 2014

\documentclass[conference]{IEEEtran}
\usepackage{graphicx}
\usepackage{times}
\usepackage{subfigure}
\usepackage{url}

@@

\title{Six ways to understand monads\\
\small{Experiences from the functional programming course at TU Delft}}

\author{\IEEEauthorblockN{Erik Meijer and Georgios Gousios}
\IEEEauthorblockA{
Software Engineering Research Group\\
Delft University of Technology\\
Delft, The Netherlands\\
Email: \{h.j.m.meijer, g.gousios\}@tudelft.nl
}
}
\maketitle

@@

  In the last few years, the popularity of functional programming as a way of
  solving computational problems has increased significantly. While most
  computer science curricula do include a course on functional programming, in
  many cases it is disconnected from practical applications, which is
  precisely where functional programming shines. To fill in this gap, we
  designed a functional programming course that  required students to
  learn by experience with real world applications. In this paper, we present
  the course's design and outline our experience from delivering it.

@@

\begin{IEEEkeywords}
functional programming, teaching
\end{IEEEkeywords}

@@

Due to a variety of reasons, including the advent of cloud computing, the rising
rate of information production and the necessity to reach the market fast,
currently, large corporations and start-ups alike are investigating alternative
programming and information storage models. As a result, during the last few
years, the practical software engineering field is witnessing a noticeable shift
towards functional programming. Scripting languages, notably Javascript and
Ruby, pioneered the introduction of functional concepts, such as closures and
lambda functions, to mainstream programming. A new wave of programming
languages, developed to overcome the expressiveness and complexity limitations
exhibited in mainstream languages, have promoted functional constructs, such as
type safe pattern matching, higher order functions and single assignment
variables, to first class citizens (Scala, {\sc c\#}).  New, functional
languages have emerged to fill in the remaining gaps (F\#, Clojure, Racket),
often introducing significant advancements in their field of specialisation
(such as Erlang in distributed fault-tolerant systems).  Finally, large scale
information processing systems such as Map/Reduce~\cite{Dean04} and domain
specific languages such as {\sc linq} have integrated functional concepts to
ease the expression of masively parallel computations.

Broadly speaking, functional programming is a style of programming in which the
primary method of computation is the application of functions to arguments.
Among other features, functional languages offer a compact notation for writing
programs, powerful abstraction methods for structuring them, and a simple
mathematical basis that supports reasoning. Many of the advanced techniques in
modern functional languages, such as monads and catamorphisms, are closely based
on principles from category theory such as functors, initial algebras, monads
and Kleisli categories.

While functional programming has been taught for long in computer science
departments~\cite{Joost93}, curricula tend to emphasize functional programming
theory rather than practical applications. Special programming languages are
used to teach functional programming specific concepts, while little connection
is made to how those concepts can be transfered to non-functional languages.
The course at TU Delft attempted to teach functional programming principles in the
classroom and then expect students to apply those concepts in the (non-purely
functional) language of their choice. The course had a very strong teaching by
example focus: students were expected to participate in both in-classroom
exercises, homework assignments and implement a real world system as a final
project.

@@

The functional programming course is elective at the MSc level. Students
from all departments in the Electrical Engineering, Mathematics and Computer
science faculty are allowed to participate. For practical reasons, the
enrollment to the course was limited to 15 participants. This permitted
us to get to know the students individually and to provide them with
personalized support.

From the students that enrolled, most have received formal introduction to
imperative and object oriented programming in their bachelors curriculum, while
through participation to other courses they had limited exposure to functional
programming (i.e. Map/Reduce in the web information systems
course). Two of the students were majoring in computer engineering, which meant
that their programming experience was restricted. Several students also had work
experience as programmers as part of their industrial placement or through
participation to start up ventures.

The diverse backgrounds of the students guaranteed that no generic
introduction to the topic would be sufficient. For this reason, we decided
to adopt a hands-on-first approach; the students would have to learn by
flexing their programming muscles, instead of being introduced to 
the theory through toy examples.

@@

Functional programming is arguably one of the oldest programming paradigms. In it is
purest form, it is based on a minimalistic theoretical background
($\lambda$-calculus). Relatively recent
formulations~\cite{Meije91, Wadle93} also introduced concepts from category
theory. While the every day use of functional programming does not necessarily
require the programmer to be aware of the theory, understanding it usually leads
to more elegant algorithmic solutions. Functional programming teaching is also
associated with advanced type systems; indeed, the flagship functional
programming languages (Haskell and the {\sc ml} family) both offer very
sophisticated support for types. Consequently, teaching the full spectrum of
functional programming theory and techniques would have been impossible in a
half-semester course; instead we decided to focus on practical aspects of data
processing, transformations and state representation using functional
techniques.

\subsection{Choosing the appropriate programming language}

Programming languages, in addition to enabling the programmer to express a
series of instructions to be executed by a computer, affect the programmer's
thought process and consequently her approach towards problem
solving~\cite{Ivers80}.  When teaching a programming course, it is important to
be able to demonstrate concepts without interference by the chosen language's
syntax. However, a non-practical language might have the opposite effect; our
experience has shown that the further a demonstration language is from practical
application, the less important students feel the taught concepts are.
Fortunately, most functional programming concepts can be expressed cleanly by
several widespread languages: for example, Javascript, Ruby and {\sc c\#} have
closures, first class and higher order functions, while a number of emerging
languages, such as Scala, Racket, {\sc f\#} and Clojure are functional languages
implemented on familiar development platforms. Apart from staying compatible
with the existing literature, there is no practical reason that necessitates
teaching strictly in a language like Haskell or {\sc ml}.

The above led us to not choose any particular language for the course.  During
the lectures, the demonstration languages where Haskell and {\sc c\#}, while at
the labs we used Scala. We actively encouraged students to use the language of
their choice on the platform of their choice to carry out homework assignments
and the course's final project.

@@

The high level goal of the course was to teach the principles of functional
programming, and the corresponding Category theoretical principles. 
More specifically, the educational purposes of the course were:

@@

  \item To introduce students to basic functional programming concepts,
    higher order functions, monads and advanced type systems.

@@

The course consisted of a series of lectures, of which two where invited
lectures by an external instructor, and laboratory sessions. In total, the
course consisted of 14 hours of lectures and 8 hours of lab work.
The students
also had to carry out a series of homework assignments and a final project.
The evaluation would be based on student performance on the final assignment. 
The lectures took place in a quarter (half semester), during 3 intense 
weeks; the students had another 6 weeks to finish their final projects.

@@

  \item Real time graph visualisation on steaming data. The
    particular example that students worked on was visualizing community
    structures for Github projects.

  \item Multi-source real time data processing. Students worked on a 
    programming language popularity index, based on data from the Github
    event stream and the StackOverflow tag cloud {\sc api}.

  \item Implementation of Haskell constructs in Javascript. Students implemented
    the Haskell Prelude (basic functions for list manipulation).

  \item Implementation of Map/Reduce algorithms in Haskell. Students installed
    and configured cloud Haskell and implemented simple Map/Reduce based
    algorithms in a distributed setting.

  \item Implementation of simple constraint solvers.

  \item Machine learning algorithms. Students implemented Na\"ive Bayes 
    classifiers and K-means clustering algorithms.

@@

To counterbalance the differences in student experience level, 
the assignments were open ended; students could drive their projects as
far as they where willing to.
The deliverables for the assignments were a repository
with the source code and short report describing the solution that the
students came up with. The students were also required to present a working
demo by the end of the course time.

@@

The purpose of the lab sessions where to help students carry out their final
assignments and further explain concepts that might not be clear to students. 
Students convened every week, presented their progress and discussed with 
each other and with the lab demonstrator design and practical issues that
arised in the implementation of their projects. To further motivate students
to work on their projects, one of the lab sessions was converted to a full
day coding sprint.

@@

The focus of teaching was to present functional programming from a practical
aspect; we were mostly interested to teach students what they can do rather than
the theory behind functional constructs. An important distinction that was made
early on was that the world is imperative; therefore while functional
programming can be a great tool for thinking about a problem, a pure functional
solutions cannot reflect on the real world. As a consequence, it is usually best
to mix programming paradigms, i.e. imperative or object-oriented programming for
state representation and functional programming for data processing. In our
experience, this distinction helped students understand that they are already
doing functional programming without noticing; enforcing data immutability to
handle multi-threading problems or passing function arguments in scripting
languages are applications of functional programming principles. The moment
students realized this fact, their attention was captured to the lectures.

During both lectures and labs, all non-whiteboard examples where actually
entered in a real programming environment (LinqPad or the Scala command line
read-evaluate loop) and the results of the evaluation where discussed with the
students. Participation was encouraged by modifying the examples and asking the
students to perform the evaluation, before the example was executed. The
examples varied from list manipulation with higher-order functions to 
a toy actor model example.

The proverbial ``six ways to understand monads'' included presentations of the
formal theoretical concepts, visualizations of monads as blenders (based on the
monadic property that once a value is captured, it is then contained),
applications in Haskell for encapsulating state (through the \textsf{Maybe}
monad), applications to generic programming in {\sc c\#}, applications to {\sc
linq} and the reactive extensions frameworks, and, finally, a whiteboard
exercise where students were prompted to participate in the ad-hoc specification
of Haskell's \textsf{IO} monad from basic principles. Once again, the examples
enabled the students to understand the composition and state containment
properties of an otherwise difficult to comprehend theoretical concept.

@@

\begin{figure*}
\centering
\label{fig:teapots}
\subfigure[Input drawing, rendered with Java graphics]{
\includegraphics[scale=0.173]{teapot-reference.png}
\label{fig:teapot-reference}
}
\subfigure[Decomposition in Scala, rendered with Java graphics]{
\includegraphics[scale=0.17]{teapot-java-canvas.png}
\label{fig:teapot-scala}
}
\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
elements]{
\includegraphics[scale=0.195]{teapot-html-divs.png}
\label{fig:teapot-html-divs}
}
\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
\includegraphics[scale=0.295]{teapot-html-canvas.png}
\label{fig:teapot-html-canvas}
}
\caption[]{Example results from the teapot exercise.}

@@

  \item Most modern programming languages can express functional programming
    constructs.

  \item Functional programming works best in data transformation scenarios.

@@

The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
graphics primitive of the student's choice using only right triangles (with one
horizontal leg).  In its
core, the exercise required students to decompose a collection of arbitrary triangles, which
comprised the input Utah Teapot model, to a collection of right triangles. The students
had to come up with the decomposition method (using some simple analytic
geometry, this was not a course on 3D graphics), a decomposition termination criterion to stop the decomposition when
triangles were too small to be rendered on screen and a method to recursively
apply the above mentioned transformations on the input data. As always, the
students could decide the implementation language of their choice. The exercise
was given as a mid-week homework assignment, between the Monday and the Friday
lectures.

The student's response was overwhelming. Even though it was made clear that the
exercise would not contribute to the final grade, the will to apply the data
processing techniques taught during the lectures motivated the students to work
very hard. The fact that they were instructed to use their favourite language
enabled them to focus on decomposing the problem in a series of testable
calculation steps rather than meddling with the intricacies of learning a new 
language. In spite of the advice, students were eager to test-drive the newly
acquired knowledge using a functional language; apart from {\sc c\#} and
Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
renderings produced by student programs can be seen in Figure~\ref{fig:teapots}.

@@

According to many students, the Map/Reduce data analysis framework was what
drove them to attend the course. The lecture on Map/Reduce featured an invited
talk by Prof. Ralf L\"ammel, who presented his formulation of Map/Reduce using
functional programming principles~\cite{Lamme08} and explained how those can be
applied to process semi-structured data. The talk resonated to students; in
their projects, students used implicit, through the selected language's library
collection methods, or explicit, by formalizing data processing as a series of
in memory Map/Reduce operations, forms of Map/Reduce to process their datasets.
Out of personal interest, one student even implemented L\"ammel's Map/Reduce
formulation in Scala, and through the language's extension mechanisms, adapted
it for use as an extension method by any Scala collection type, including
parallel collections.

@@

\section*{Acknowledgements and Availability}
We would like to thank the course participants for their enthusiasm and
dedication. The source code to all student projects and the teapot
exercise can be found at (or linked from) \url{http://swerl.tudelft.nl/bin/view/Main/FunctionalProgrammingCourse}. This work is partially supported by
Marie Curie {\sc ief} 298930 -- {\sc sefunc}.

\bibliography{paper}
\bibliographystyle{ieeetr}
\end{document}

>>>>

\documentclass[conference]{sig-alternate}
\usepackage{graphicx}
\usepackage{times}
\usepackage{subfigure}
\usepackage{hyperref}

@@

%\author{\IEEEauthorblockN{Erik Meijer and Georgios Gousios}
%\IEEEauthorblockA{
%Software Engineering Research Group\\
%Delft University of Technology\\
%Delft, The Netherlands\\
%Email: \{h.j.m.meijer, g.gousios\}@tudelft.nl
%}
%}

\title{Teaching Applied Functional Programming\\
\small{Experiences from the functional programming course at TU Delft}
}

\numberofauthors{2}

\author{
\alignauthor
Erik Meijer\\
       \affaddr{Delft University of Technology}\\
       \affaddr{Delft, The Netherlands}\\
       \email{h.j.m.meijer@tudelft.nl}
\alignauthor
Georgios Gousios\\
       \affaddr{Delft University of Technology}\\
       \affaddr{Delft, The Netherlands}\\       
       \email{g.gousios@tudelft.nl}
}

\maketitle

@@

  Recently, the popularity of functional programming as a way of
  solving computational problems has increased significantly. While most
  computer science curricula do include a course on functional programming, in
  many cases it is disconnected from practical applications, which is
  precisely where functional programming shines. To fill in this gap, we
  designed a functional programming course that required students to
  learn by experience in real world applications. In this paper, we present
  the course's design and outline our experience from delivering it.

@@

\category{K.3.2}{Computing Milieux}{Computers and Education}[Computer and Information Science Education]
\category{D.1.1}{Software}{Programming Techniques}[Applicative (Functional) Programming]

\terms{Languages}

\keywords{functional programming, teaching}

@@

Due to a variety of reasons, including the advent of cloud computing, the rising
rate of information production and the necessity to reach the market fast,
currently, large corporations and start-ups alike are investigating alternative
programming and information storage models. As a result, during the last few
years, the leading edge of practical software engineering field is witnessing a
slight but noticeable shift towards functional programming ({\sc
fp}).\footnote{See, for example, the reactive manifesto:
\href{http://www.reactivemanifesto.org/}{http://www.reactivemanifesto.org/}}
Scripting languages, notably Javascript and Ruby, pioneered the introduction of
functional concepts, such as closures and lambda functions, to mainstream
programming. A new wave of programming languages, developed to overcome the
expressiveness and complexity limitations exhibited in mainstream languages,
have promoted functional constructs, such as type safe pattern matching, higher
order functions and single assignment variables, to first class citizens (Scala,
{\sc c\#}).  New, functional languages have emerged to fill in the remaining
gaps (F\#, Clojure, Racket), often introducing significant advancements in their
field of specialisation (such as Erlang in distributed fault-tolerant systems).
Finally, large scale information processing systems such as
Map/Reduce~\cite{Dean04} and reactive programming~\cite{Meije12} and domain
specific languages such as {\sc linq}~\cite{Meije11} have integrated functional
concepts to ease the expression of massively parallel computations.

Strictly speaking, {\sc fp} is a style of programming in which the primary
method of computation is the application of functions to arguments.  Among other
features, functional languages offer a compact notation for writing programs,
powerful abstraction methods for structuring them, and a simple mathematical
basis that supports reasoning. Many of the advanced techniques in modern
functional languages, such as monads~\cite{Wadle93} and
catamorphisms~\cite{Meije91}, are closely based on principles from category
theory such as functors, initial algebras, monads and Kleisli categories.

While {\sc fp} has been taught for long in computer science
departments~\cite{Joost93}, curricula tend to emphasize {\sc fp} theory rather than practical applications. Special programming languages are
used to teach {\sc fp} specific concepts, while little connection
is made to how those concepts can be transfered to solving practical software
engineering problems.

The presented course runs at TU Delft for two consecutive years. Our goal, and
the course's moto, is to teach students to  ``think like ({\sc fp})
fundamentalists and code like hackers''. The method we follow is teaching
{\sc fp} principles in the classroom and then expecting students
to apply those concepts in challenging practical projects. The course has a very
strong teaching by example focus: students are expected to participate in both
in-classroom exercises, homework assignments and implement a real world system
as a final project, all within a half-semester (6 -- 8 weeks).

@@

The {\sc fp} course is elective at the MSc level. Students from
all departments in the Electrical Engineering, Mathematics and Computer science
faculty are allowed to participate. In the first year, we restricted the
enrollment to the course to 15 participants; we lifted this
restriction during the second year, as we found that the format could allow for
more students to participate.

From the students that enroll, most have received formal introduction to
imperative and object oriented programming in their bachelors curriculum, while
through participation to other courses they had limited exposure to {\sc fp}
(i.e. Map/Reduce in the Web Information Systems course). Some students are
majoring in computer engineering, which meant that their programming experience
was more restricted. Most students also have work experience as programmers as
part of their industrial placement or through participation to start up
ventures.

The diverse backgrounds of the students guarantees that no ge\-ne\-ric
introduction to the topic would be sufficient. For this reason, we decided
to adopt a hands-on-first approach; the students would have to learn by
flexing their programming muscles, instead of being gently introduced to 
the theory through toy examples. During the lectures, we made clear to the
students that it is the delta in learning that we expect them to
demonstrate rather than impressive technical achievements.

@@

{\sc fp} is arguably one of the oldest programming paradigms. In
it is purest form, it is based on a minimalistic theoretical background
($\lambda$-calculus). Relatively recent formulations~\cite{Meije91, Wadle93}
also introduced concepts from category theory. While the every day use of
{\sc fp} does not necessarily require the programmer to be aware
of the theory, understanding it usually leads to more elegant algorithmic
solutions. Teaching {\sc fp} is also associated with advanced type
systems; indeed, the flagship {\sc fp} languages (Haskell, Scala) both offer very sophisticated support for types. Consequently,
teaching the full spectrum of {\sc fp} theory and techniques is impossible in a half-semester course; instead we focus on
practical aspects of data processing, transformations and state representation
using functional techniques.

\subsection{Choosing a programming lan\-guage}

Programming languages, in addition to enabling the programmer to express a
series of instructions to be executed by a computer, affect the programmer's
thought process and consequently her approach towards problem
solving~\cite{Ivers80}.  When teaching a programming course, it is important to
be able to demonstrate concepts without interference by the chosen language's
syntax. However, a non-practical language might have the opposite effect; our
experience has shown that the further a demonstration language is from practical
application, the less important students feel the taught concepts are.
Fortunately, most {\sc fp} concepts can be expressed cleanly by
several widespread languages: for example, Ja\-va\-script, Ruby and {\sc c\#} have
closures, first class and higher order functions, while a number of emerging
languages, such as Scala, Racket, {\sc f\#} and Clojure are functional languages
implemented on familiar development platforms. Apart from staying compatible
with the existing literature, there is no practical reason that necessitates
teaching in a strict {\sc fp} language like Haskell or {\sc ml}.

The above led us to not choose any particular language for the course.  During
the lectures, the demonstration languages where Haskell, {\sc c\#} and {\sc php}, while at the labs we used Scala. We actively encouraged students to use the language of
their choice on the platform of their choice to carry out homework assignments
and the course's final project.

@@

The high level goal of the course was to teach the principles of {\sc fp}, and the corresponding Category theoretical principles. More specifically, the educational purposes of the course were:

@@

  \item To introduce students to basic {\sc fp} concepts,
    higher order functions, monads and advanced type systems.

@@

The course consists of a series of lectures. In total, the
course consisted of 15 hours of lectures. 
The students
also have to carry out a series of optional homework assignments and a final project.
The evaluation is based on student performance on the final assignment. 
The lectures take place in a quarter (half semester), during 6 intense 
weeks; the students have another 2-3 weeks to finish their final projects.

@@

  \item Real time graph visualisation on steaming data.

  \item Haskell constructs (e.g. the Prelude or abstract data types) in Javascript or even {\sc php}.

  \item Interpreters and byte code converters between languages.

  \item Simple constraint solvers and machine learning algorithms.

  \item Asynchronous request processing in network servers

@@

To counterbalance the differences in student experience level, the assignments
are open ended; students can drive their projects as far as they are willing to,
as long as the experience is pleasant enough to allow them to learn. The deliverables for the
assignments are a repository with the source code and short report describing
the solution that the students came up with. The students are also required to
present a working demo by the end of the course time.

@@

In the first year it was offered, the course included lab sessions.  The purpose
of the lab sessions were to help students carry out their final assignments and
further explain concepts that might not be clear to them.  Students convened
every week, presented their progress and discussed with each other and with the
lab demonstrator design and practical issues that hindered the implementation of
their projects. To further motivate students to work on their projects, one of
the lab sessions was converted to a full day coding sprint. However, we found
out that participation was diminishing and the sessions were not of much interest to the students,
who preferred to have more time to work on the projects. We therefore canceled
the lab sessions in the course's second year.

\subsection{Organizing the work}

The higher number of students in the second year of the project made the
organization of student work challenging. To better organize and observe the
student's work, we made active use of Github. Github is a code hosting site that
includes social networking features (such as notifications on events, following
of projects etc) and a dashboard to observe them. We created an organization for
the course\footnote{\href{https://github.com/organizations/fptudelft}{https://github.com/organizations/fptudelft}} and in it we
created repositories for the projects shared among all students registered on
the course. We also created a specialized repository for answering general
questions about the course, using the repository's issue tracking facility.

Github's notification functionality enabled us and the teaching assistants to
answer questions about the student projects in almost real time. We could also
observe work done by the students while it was done and this allowed us to
formulate an early opinion on their progress. We then informally discussed this with the students during idle time (e.g. lecture breaks), giving two types of feedback: praise to students who were doing good, help and support to students that were struggling.

@@

The focus of teaching was to present {\sc fp} from a practical
aspect; we were mostly interested to teach students what they can do rather than
the theory behind functional constructs. An important distinction that was made
early on was that the world is imperative; therefore while {\sc fp} 
 can be a great tool for thinking about a problem, a pure functional
solutions cannot reflect on the real world. As a consequence, it is usually best
to mix programming paradigms, i.e. imperative or object-oriented programming for
state representation and {\sc fp} for data processing. In our
experience, this distinction helped students understand that they are already
doing {\sc fp} without noticing; enforcing data immutability to
handle multi-threading problems or passing function arguments in scripting
languages are applications of {\sc fp} principles. The moment
students realized this, their attention was captured to the lectures.

During both lectures and labs, all non-whiteboard examples where actually
entered in a real programming environment (LinqPad in {\sc c\#} or the Scala
command line read-evaluate loop) and the results of the evaluation where
discussed with the students. Participation was encouraged by modifying the
examples and asking the students to perform the evaluation, before the example
was executed.
The examples varied from list manipulation with higher-order
functions to a toy actor model example.
Moreover, we made extensive use of the whiteboard to demonstrate abstract
concepts like type inference and combinators. Typically, we would
write a function signature in Haskell and then expected students to come up 
with its implementation and vice versa.

Of particular interest to the students was the presentation of monads. The
students were teased that they will learn monads in no less than 6 different
ways. Those included presentations of the formal theoretical concepts,
visualizations of monads as blenders (based on the monadic property that once a
value is captured, it is then contained), applications in Haskell for
encapsulating state (through the \textsf{Maybe} monad), applications to generic
programming in {\sc c\#}, applications to {\sc linq} and the reactive extensions
frameworks, and, finally, a whiteboard exercise where students were prompted to
participate in the ad-hoc specification of Haskell's \textsf{IO} monad from
basic principles. Once again, the examples enabled the students to understand
the composition and state containment properties of an otherwise difficult to
comprehend theoretical concept.

@@

\begin{figure*}
\centering
\label{fig:teapots}
\subfigure[Input model]{
\includegraphics[scale=0.14]{reference.png}
\label{fig:teapot-reference}
}
\subfigure[Decomposed model]{
\includegraphics[scale=0.20]{decomposed.png}
\label{fig:teapot-scala}
}
\subfigure[Detail: The teapot's handle]{
\includegraphics[scale=0.20]{detail.png}
\label{fig:teapot-scala}
}

\caption[]{The teapot input and decomposition results}

@@

  \item Most modern programming languages can express {\sc fp} constructs.

  \item {\sc fp} works best in data transformation scenarios.

@@

The first time the course was given, the exercise consisted of rendering the
Utah Teapot~\cite{Torre06} using any graphics primitive of the student's choice
using only right triangles with one horizontal leg (see
Figure~\ref{fig:teapots}). In its core, the exercise required students to
decompose a collection of arbitrary triangles, which comprised the input Utah
Teapot model, to a collection of right triangles~\cite{Beckm13}. The students
had to come up with the decomposition method (using analytical geometry), a decomposition termination criterion to
stop the decomposition when triangles were too small to be rendered on screen
and a method to recursively apply the above mentioned transformations on the
input data. As always, the students could decide the implementation language of
their choice. The exercise was given as a mid-week homework assignment, between
the Monday and the Friday lectures.

During the second run of the course, the exercise was optimizing decomposition
rather than implementing it. We provided the students with a basic implementation in {\sc php} and asked them to optimize the code by applying new algorithmic techniques and taking into account new language features implemented in a 
pre-release version of Facebook's {\sc php} language version (codenamed Hack).
To further motivate them, we converted the exercise to a programming
contest with prizes for the winning team.

Both times, the students' response was overwhelming. Even though it was made
clear that the exercise would not contribute to the final grade, the will to
apply the data processing techniques taught during the lectures motivated the
students to work very hard. In the first year, the fact that they were
instructed to use their favourite language enabled them to focus on decomposing
the problem in a series of testable calculation steps rather than meddling with
the intricacies of learning a new language. In spite of the advice, students
were eager to test-drive the newly acquired knowledge using a functional
language; apart from {\sc c\#} and Javascript, solutions were also provided in
Scala, Scheme and Haskell. In the second year, the students learned the
value of clean design and breaking code changes to obtain massive
optimizations~\cite{Gousi13}.

@@

According to many students, the Map/Reduce data analysis framework was what
drove them to attend the course. In the first year, the lecture on Map/Reduce featured an invited
talk by Prof. Ralf L\"ammel, who presented his formulation of Map/Reduce using
functional programming principles~\cite{Lamme08} and explained how those can be
applied to process semi-structured data. The talk resonated to students; in
their projects, students used implicit, through the selected language's library
collection methods, or explicit, by formalizing data processing as a series of
in memory Map/Reduce operations, forms of Map/Reduce to process their datasets.
Out of personal interest, one student even implemented L\"ammel's Map/Reduce
formulation in Scala, and through the language's extension mechanisms, adapted
it for use as an extension method by any Scala collection type, including
parallel collections.

@@

\section*{Acknowledgements and Availability}
We would like to thank the course participants for their enthusiasm and
dedication. The source code to all student projects and the teapot
exercise can be found at (or linked from) \href{https://github.com/organizations/fptudelft}{https://\-github.com\-/organizations\-/fptudelft}. This work is partially supported by Marie Curie {\sc ief} 298930 -- {\sc sefunc}.

\bibliography{paper}
\bibliographystyle{abbrv}
\end{document}
